---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Setting options for the overall file
```{r setoptions, echo=TRUE}
library(knitr)
opts_chunk$set(echo = TRUE, results = "asis")
```
## Loading the data

Load the data from CSV to variable activity
```{r loaddataset}
activity = read.csv("activity.csv")
str(activity)
```

##Preprocessing the Data

```{r preprocess}
library(lubridate)
activity$date <- ymd(activity$date)
str(activity)
summary(activity)
```


## What is mean total number of steps taken per day?
```{r meansteps}
library(dplyr)
steptakenbyday <- activity %>% 
                    na.omit() %>% 
                         group_by(date) %>% 
                              summarise(totalsteps = sum(steps)) 

```

Verifying filtered/aggregated data
```{r}
head(steptakenbyday, n=10)
str(steptakenbyday)
```

## Make Historgram of total number of steps taken each day

```{r}
library(ggplot2)
qplot(totalsteps, data=steptakenbyday, geom="histogram", main = "Total number of steps taken each day", xlab = "Total Number of steps", ylab="Frequency(Count of Days)", binwidth=2000)
stepsByDayCentralMeasures <- steptakenbyday %>% 
                    summarise(meanStepsByDay=as.numeric(mean(totalsteps)), 
                         medianStepsByDay=as.numeric(median(totalsteps)))

meanst <- mean(steptakenbyday$totalsteps)
medianst <- median(steptakenbyday$totalsteps)
```

The mean of the total number of steps taken per day is `r meanst`  
The median of the total number of steps taken per day is `r medianst`

## What is the average daily activity pattern?
```{r meandailyactivity}
avgdailyactivity <- activity %>% 
                         na.omit() %>% 
                              group_by(interval) %>% 
                                   summarise(avgsteps = mean(steps))

head(avgdailyactivity, 10)


```

Time Series Plot 

```{r plot}
qplot(x=interval, y=avgsteps, data=avgdailyactivity, geom="line", main = "Average number of steps taken each day by Interval", xlab = "Average Number of steps", ylab="Frequency(5 Minute Interval)")
```


```{r}
which.max(avgdailyactivity$avgsteps)
maxindex <- which.max(avgdailyactivity$avgsteps)
intervalav <- avgdailyactivity[maxindex,1]
avsteps <- avgdailyactivity[maxindex,2]

```
Max steps 5 minute interval is `r avsteps` for interval `r intervalav`


## Imputing missing values

Identify missing values
```{r}
summiss<- sum(is.na(activity$steps))
```
Count of missing values in Dataset is `r summiss`

Startegy for filling missing value in dataset
```{r}
nadataset <- activity[which(is.na(activity$steps)),]
nadataset <- nadataset %>%
               inner_join(avgdailyactivity, by="interval") %>%
                    mutate(steps=avgsteps) %>%
                         select(-avgsteps)
head(nadataset)
```

Combined new dataset with default missing value
```{r}
combineddataset  <- activity

combineddataset[which(is.na(combineddataset$steps)), 1] <- nadataset[ , 1]

head(combineddataset)
sum(!complete.cases(combineddataset))

newStepsByDay <- combineddataset %>% 
                     group_by(date) %>% 
                         summarise(totalSteps=sum(steps))

head(newStepsByDay)
```
histogram on new dataset
```{r}
qplot(totalSteps, data = newStepsByDay, geom = "histogram", main = "Total number of steps taken each day with default value for Steps = NA", xlab = "Total Number of steps", ylab="Frequency(Count of Days)", binwidth=2000)
tapply(nadataset$steps, as.factor(nadataset$date), sum)

newStepsByDayCentralMeasures <- newStepsByDay %>% 
               summarise(meanStepsByDay=as.numeric(mean(totalSteps)), 
               medianStepsByDay=as.numeric(median(totalSteps)))
rbind(stepsByDayCentralMeasures, newStepsByDayCentralMeasures)


```


## Are there differences in activity patterns between weekdays and weekends?

```{r}
combineddataset <- combineddataset %>% 
               mutate(weekdays=ifelse(weekdays(date) == "Saturday" | 
                                      weekdays(date) == "Sunday", 
                                      "weekend", 
                                      "weekday"))
combineddataset$weekdays <- as.factor(combineddataset$weekdays)
str(combineddataset)

revisedStepsByInterval <- combineddataset %>%
                          group_by(interval, weekdays) %>%
                              summarise(averageSteps=mean(steps))
head(revisedStepsByInterval, n=10)

qplot(x=interval,
      y=averageSteps,
      data=revisedStepsByInterval,
      geom="line",
      main="Average number of steps taken for each Interval, across all days\n",
      xlab="\nInterval",
      ylab="Number of Steps\n",
      facets=weekdays ~ .)



```

Pattern does not differ much between weekday and weekend